---
title: "Beginners' R -- Plotting"
author: "Katherine Tansey"
date: "27 November 2017"
output:
  html_document:
    depth: 4
    keep_md: yes
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---


# Let's make a plot 

First, we would need to load the *ggplot2* library. However, if we have loaded the *tidyverse* library, then *ggplot2* is already loaded. 

This is because *tidyverse* package is a package of packages, does not contain its own functions per se but includes a [variety of other packages](https://blog.rstudio.org/2016/09/15/tidyverse-1-0-0/), like [dplyr](http://dplyr.tidyverse.org/), [tidyr](http://tidyr.tidyverse.org/), [ggplot2](http://ggplot2.org/). Therefore by installing and loading the *tidyverse* package, we install and load all of these other packages as well.  

Using the diamonds dataset:

1. Examine the diamonds dataset

```{r}

```

2. Make a scatter plot using any two *numeric* columns in the diamonds dataset.      
*HINT* [dbl and int](http://uc-r.github.io/integer_double/) are indicators of numeric columns       
*REMEMBER* basic syntax for a scatter plot in ggplot is this:

```{r, eval = FALSE}
ggplot(data = mtcars, mapping = aes(x=mpg, y=hp)) +
	geom_point()
```

```{r}

```

3. What happens if you do not include *data = * or *mapping = *?

```{r}

```

4. What happens if you do not include the line starting with geom (just run the first line)?

```{r}

```


# Attributes and aesthetics 

Using the diamonds dataset:    

1. Remake the scatter plot you made above, but assign a *non-numeric* variable to **color**.    

```{r}

```

2. Remake the scatter plot above, but now assign a *non-numeric* variable to **shape** and set the **color** of all points to be red.     

```{r}

```

3. Remake the scatter plot above, but now assign a *numeric* variable to **size** and a *non-numeric* variable to **color**. Change the **shape** of all the points to be a filled-in square (shape = 15).  

```{r}

```

4. Remake the scatter plot above.
What happens if you set **color** as an aesthetic (assign a *non-numeric* variable to color) *AND* set a **color** (red) for all points in the geom_point command?

```{r}

```

Why do you think this is happening?   


# Geometries 

#  One Variable -- Continuous

Using the diamonds dataset:

1. Make a histogram of the carat variable.

```{r}

```

2. Experiment with binwidth.     
Add *binwidth = #* in the parentheses for *geom_histogram*. Play around with the number (#).   

```{r}

```

What is binwidth doing?   
Does the plot and interpretation become different as you alter the binwidth?  

3. Change the histogram to a density plot.   

```{r}

```

4. Alter the density plot code above to add an aesthetic for **color** to be the variable clarity (color = color). 

```{r}

```

5. Alter the density plot code above, but change the word color to **fill**. 

```{r}

```

6. Does the **color** and **fill** variables work the same for *geom_histogram*?       
Try it out.

```{r}

```

# One Variable -- Discrete

Make a bar chart using the diamonds dataset. Select a *non-numeric* variable to chart.    

```{r}

``` 


# Two Variable -- Both continuous 

Using the gapminder data:          

1. Make a scatter plot of life expectancy by GDP

```{r}

``` 

2. Alter the plot above to use the log of GDP (*log(gdpPercap)*).      
[See here](http://astrostatistics.psu.edu/su07/R/html/base/html/Log.html) for how to take log of a variable      

```{r}

``` 

3. Alter the plot above to **color** the points by continent

```{r}

``` 

4. Alter the plot above to change the **size** of the points by the population

```{r}

``` 

5. There are a lot of points on the plot, so let's clean up the plot a bit and just focus on one single year, the latest year in the data 2007.   

Using dplyr:     
We can filter the data to only include rows where year is 2007. We can do this two ways:   
    1. Filter gapminder data within the ggplot2 command    
    2. Filter the gapminder data and save the out as a new data frame, then read this data frame in as the data for ggplot2 command    
 
```{r}

``` 

6. Turn off legends.   
[See here](http://www.cookbook-r.com/Graphs/Legends_(ggplot2)/) for more information about altering legends. Adding *theme(legend.position="none")* as a new line will remove the legend.   
**REMEMBER** lines have to be connected by a +     

```{r}

``` 

7. Add to the plot a line of best fit.  
ggplot2 allows for multiple similar geoms to be plotted onto the same plotting window (remember the idea of layering).  
Like above, you add a new layer to the plot but using the **+** and adding the additional geoms/layers.         
We want to run a linear smooth model (*method = "lm"*) and turn off the se (*se = FALSE*), so add this code to the plot *geom_smooth(method = "lm", se = FALSE)* 

```{r}

``` 

geom_smooth is creating a line for *every* continent, why do you think this is?    
Think about how inheritance of the aesthetics works.     

We want to look at a single line, a *global trend* rather than continental trends.      
Each geom can have their own local aesthetic set. If we move **color** and **size** from the top line into geom_point as aesthetics, we will now plot the global trend line with geom_smooth.   

The code for geom_point should now look like this: *geom_point(aes(color = continent, size = pop))*      
Try to alter the code to create this.       

```{r}

``` 

# Over plotting    

Using the gapminder data:

1. Remake the scatter plot of life expectancy by GDP using all the gapminder data.       

```{r}

```

2. Change geom_point to geom_jitter.   
Within the geom_jitter command, include arguments for alpha (*alpha = #*). 

```{r}

```   

*What do you think the alpha option is doing?*    

3. Change the above code to include **shape** with geom_jitter. Change the shape to a hollow circle (shape = 1).    

```{r}

```

# Two Variable -- Continuous and Time

Use the gapminder data:

1. Create a plot showing the change in population over time for the United Kingdom.     
We need to filter the dataset to just include the country United Kingdom. Like above there are two ways to do this:
    1. Filter gapminder data within the ggplot2 command   
    2. Filter the gapminder data and save the out as a new data frame, then read this data frame in as the data for ggplot2 command    
    
```{r}

``` 

2. If you wanted to compare the UK change in population to France, how could you do that? 

```{r}

``` 

# Two Variable -- Discrete and Continuous 

Using gapminder dataset:

1. Make a box plot of life expectancy by continent.  
*continent* need to wrapped in the command *factor()*, as was done in the example code above.     

```{r}

``` 
*What does a box in the plot indicate?*

2. Using the above plot, add points for each data point to the plot. 
*Should we be using some overplotting reduction techniques?* 

```{r}

``` 

You can tell jitter a range to jitter within, [see here](http://ggplot2.tidyverse.org/reference/position_jitter.html). Adding position information to the geom_jitter command can alter the spread of the points.  

3. Add *position = position_jitter(width = 0.1, height = 0.1)* within the parentheses for geom_jitter to the plot made above, does this look nicer?

```{r}

``` 

4. Change the above plot into a violin plot.   

```{r}

``` 

*Which one do you think is more informative, the box plot or violin plot?*


# Facet
    
Using the gapminder data:

1. Make a box plot for life expectancy by year faceted for continent (make multiple plots in rows). 

```{r}

``` 

2. Make the boxplot for life expectancy by continent faceted for year (make multiple plots in rows).    

```{r}

```  

Using the diamonds data set:      

3. Make a scatter plot of price vs carat and facet it by cut.    
*REMEMBER* the over plotting techniques.    

```{r}

``` 

# Coordinates

Using the mtcars dataset:

1. Make a scatter plot of mpg vs hp. Add a smoothing line (use the default settings).  

```{r}

```

2. Using the plot above, alter the x axis using the command [xlim](http://ggplot2.tidyverse.org/reference/lims.html).      
xlim is another layer that is added to the plot and is included as its own line connected to the rest of the plot code by a **+**    
Change the x axis to range from 10 to 20

```{r}

```

3. Alter the x axis using [coord_cartesian](http://ggplot2.tidyverse.org/reference/coord_cartesian.html) instead.    
Change the x-axis to range from 10 to 20.     
Because coord_cartesian takes both x and y axis information, you need to tell it which axis you want to alter. The code within the coord_cartesian command should look something like this : xlim=c(7.5,10)    (except we want different numbers!)

```{r}

```

# Themes

Using the gapminder data:

1. Remake a line plot for the population by year for the UK. 
*REMEMBER* need to filter the gapminder data for only the rows where country is the United Kingdom

```{r}

```

2. Add *theme_bw()* to the plot.    
This is another layer being add, so the *theme_bw()* line of code is add to the plot with a **+** connecting it to the rest of the code.    

```{r}

```

3. **Challenge**. Get the total population for each continent for each year, and the create a line plot showing the change in population for each continent over time. Change the theme to be theme_bw. 

```{r}

```
   
